---
title: "02: General Wnt/PCP interactome description"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

# Remove all objects from the workspace

```{r clear-workspace, message=FALSE, warning=FALSE}
rm(list = ls())
```

# Libraries

```{r libraries, message=FALSE, warning=FALSE}
library(dplyr)
library(here)
library(DEP)
library(factoextra)
library(tibble)
library(tidyr)
library(UpSetR)
```

# Data input

```{r data-input, message=FALSE, warning=FALSE}
load(here('outputs',  '01_DEP-processing', '01_DEP-df-wide.RData'))
load(here('outputs', '01_DEP-processing', '01_DEP-results-table.RData'))
```

Prepare a folder for the outputs
```{r outputs-folder, message=FALSE, warning=FALSE}
#dir.create(here('outputs', '02_PCP-interactome-description'))
```


# Perform a PCA on samples

```{r PCA, message=FALSE, warning=FALSE}
tmp <- df_wide %>%
  select(name, CTRL_1:DVL3_6) %>%
  rowwise() %>%
  mutate(most_variable = sd(c_across(CTRL_1:DVL3_6))) %>%
  arrange(desc(most_variable)) %>%
  ungroup() %>%
  slice_head(n = 500) %>%
  select(-most_variable) %>%
  column_to_rownames(var = "name") %>%
  t() %>%
  as.data.frame() %>%
  mutate(condition = c("CTRL", "CTRL", "ROR1", "ROR1", "ROR2", "ROR2", "PRICKLE1", "PRICKLE1",
                "VANGL2", "VANGL2", "DVL3", "DVL3",
                "CTRL", "CTRL", "ROR1", "ROR1", "ROR2", "ROR2", "PRICKLE1", "PRICKLE1", 
                "VANGL2", "VANGL2", "DVL3", "DVL3",
                "CTRL", "CTRL", "ROR1", "ROR1", "ROR2", "ROR2", "PRICKLE1", "PRICKLE1", 
                "VANGL2", "VANGL2", "DVL3", "DVL3")) 

 #svg(here('outputs', '02_PCP-interactome-description', '02_PCA-conditions.svg'))
 prcomp(tmp[, -501], center = TRUE, scale = FALSE) %>%
    fviz_pca_ind(habillage = tmp$condition, palette = c("#8491B4FF", "4DBBD5FF", "#00A087FF", "#3C5488FF", "#F39B7FFF",  "#E64B35FF")) 
# dev.off()
```

# Number of upregulated proteins per bait

Find which proteins are upregulated based on threshold of logFC > 1 and adjusted p-value < 0.05
```{r upregulated-proteins, message=FALSE, warning=FALSE}
data_results <- data_results %>%
  mutate(up_DVL3 = ifelse(DVL3_vs_CTRL_p.adj < 0.05 & DVL3_vs_CTRL_ratio > 1, 1, 0)) %>%
  mutate(up_ROR1 = ifelse(ROR1_vs_CTRL_p.adj < 0.05 & ROR1_vs_CTRL_ratio > 1, 1, 0)) %>%
  mutate(up_ROR2 = ifelse(ROR2_vs_CTRL_p.adj < 0.05 & ROR2_vs_CTRL_ratio > 1, 1, 0)) %>%
  mutate(up_PRICKLE1 = ifelse(PRICKLE1_vs_CTRL_p.adj < 0.05 & PRICKLE1_vs_CTRL_ratio > 1, 1, 0)) %>%
  mutate(up_VANGL2 = ifelse(VANGL2_vs_CTRL_p.adj < 0.05 & VANGL2_vs_CTRL_ratio > 1, 1, 0))
```

Plot the barplot of upregulated proteins per bait
```{r upregulated-proteins-per-bait, message=FALSE, warning=FALSE}
#svg(here('outputs', '02_PCP-interactome-description', '02_barplot-upregulated-preys-per-bait.svg'))
data_results %>%
  select(name, starts_with("up")) %>%
  pivot_longer(cols = up_DVL3:up_VANGL2, names_to = "bait", values_to = "upregulated") %>%
  mutate(bait = gsub("up_", "", bait)) %>%
  group_by(bait) %>%
  summarise(n_upregulated = sum(upregulated)) %>%
  ggplot(aes(x = reorder(bait, -n_upregulated), y = n_upregulated, fill = bait)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  theme_minimal(base_size = 17)+
  labs(x = "Bait", y = "Number of upregulated preys") +
  scale_fill_manual(values = c( "4DBBD5FF", "#00A087FF", "#3C5488FF", "#F39B7FFF",  "#E64B35FF"))
#dev.off()
```

# UpSet plot of upregulated proteins

UpSet plot of upregulated preys in particular baits.
```{r  upset-plot, message=FALSE, warning=FALSE}
#svg(here('outputs', '02_PCP-interactome-description', '02_upsetPlot-upregulated-preys.svg'))
data_results %>%
  select(starts_with("up_")) %>%
  rename_with(~ sub("up_", "", .x), everything()) %>%
  upset(., order.by = "freq")
#dev.off()
```

# Human Cell Map localization

In order to determine whether interactions of particular bait happen in a specific cellular compartment, upregulated preys were mapped to the human cell map experimental data.

```{r human-cell-map, message=FALSE, warning=FALSE}
# Data input for Human Cell Map:
hcm <- read.delim(here('data', 'preys-latest.txt'))

# Map onto the Human Cell Map
data.hcm <- data_results %>%
  left_join(., hcm %>% select(symbol, MMF.localization), by = c("name" = "symbol"))

data.hcm[is.na(data.hcm)] <- "-" #If protein was not mapped, use "-" instead

# Split the MMF.localization by "," in case there are multiple localizations mapped to the same protein
data.hcm <-  data.hcm %>%   
  mutate(NMF.loc = strsplit(as.character(MMF.localization), ", ")) %>% 
  unnest(NMF.loc)

# Create long table
data.hcm <- data.hcm %>%
  select(name, starts_with("up_"), NMF.loc) %>%
  rename_with(~ sub("up_", "", .x), everything()) %>%
  pivot_longer(cols = DVL3:VANGL2,
               names_to = "bait",
               values_to = "significant")

#svg(here('outputs', '02_PCP-interactome-description', '02_facetBarplot-humancellmap_multicolor.svg'))
data.hcm %>%
  group_by(bait, NMF.loc) %>%
  summarise(n_preys = sum(significant)) %>%
  filter(n_preys != 0) %>%
  mutate(top_loc = case_when(
    n_preys == max(n_preys) ~ "max",
    TRUE ~ "")
  ) %>%
  mutate(top_localization = case_when(
    top_loc == "max" & bait == "DVL3" ~ "DVL3",
    top_loc == "max" & bait == "ROR1" ~ "ROR1",
    top_loc == "max" & bait == "ROR2" ~ "ROR2",
    top_loc == "max" & bait == "PRICKLE1" ~ "PRICKLE1",
    top_loc == "max" & bait == "VANGL2" ~ "VANGL2",
    TRUE ~ ""
      )) %>%
  ggplot(aes(x = NMF.loc, y = n_preys, fill = top_localization)) +
  geom_bar(stat = "identity")+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  labs(x = "Prey localization in Human Cell Map",
       y = "Number of preys") + 
  facet_grid(bait ~ .) + 
  scale_fill_manual(values = c("#b8B8B8", "4DBBD5FF", "#00A087FF", "#3C5488FF", "#F39B7FFF","#E64B35FF" ))
#dev.off()


```

