---
title: "03: REPRINT analysis"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

# Libraries

```{r libraries, message=FALSE, warning=FALSE}
library(dplyr)
library(here)
library(DEP)
```

# Data input

```{r data-input, message=FALSE, warning=FALSE}
load(here('outputs', '01_DEP-processing', '01_DEP-df-wide.RData'))
```

Prepare a folder for the outputs
```{r outputs-folder, message=FALSE, warning=FALSE}
#dir.create(here('outputs', '03_REPRINT-analysis'))
if (!dir.exists(here('outputs', '03_REPRINT-analysis'))){
  dir.create(here('outputs', '03_REPRINT-analysis'))
}
```

# Filter relevant columns

```{r data-filtering, message=FALSE, warning=FALSE}
# Filter protein ID, gene name, length of protein sequence, and MS/MS counts
data <- df_wide %>%
  select(ID, name, Sequence.length,
         starts_with("MS.MS.count."))

# Renaming columns
data <- data %>%
  rename('PROTID' = 'ID',
         'GENEID' = 'name',
         'PROTEIN' = 'Sequence.length',
         'CTRL_01_NUMSPECSTOT' = 'MS.MS.count.1R.EE.01' ,
         'CTRL_02_NUMSPECSTOT' =	'MS.MS.count.1R.EE.02' ,
         'ROR1_01_NUMSPECSTOT' = 'MS.MS.count.1R.EE.03' ,
         'ROR1_02_NUMSPECSTOT' =	'MS.MS.count.1R.EE.04' ,
         'ROR2_01_NUMSPECSTOT' = 'MS.MS.count.1R.EE.05' ,
         'ROR2_02_NUMSPECSTOT' = 'MS.MS.count.1R.EE.06' ,
         'PRICKLE1_01_NUMSPECSTOT' = 'MS.MS.count.1R.EE.07' ,
         'PRICKLE1_02_NUMSPECSTOT' = 'MS.MS.count.1R.EE.08' ,
         'VANGL2_01_NUMSPECSTOT' = 'MS.MS.count.1R.EE.09' ,
         'VANGL2_02_NUMSPECSTOT' = 'MS.MS.count.1R.EE.10' ,
         'DVL3_01_NUMSPECSTOT' = 'MS.MS.count.1R.EE.11' ,
         'DVL3_02_NUMSPECSTOT' =	'MS.MS.count.1R.EE.12' ,
         'CTRL_03_NUMSPECSTOT' =	'MS.MS.count.2R.EE.01' ,
         'CTRL_04_NUMSPECSTOT' =	'MS.MS.count.2R.EE.02' ,
         'ROR1_03_NUMSPECSTOT' =	'MS.MS.count.2R.EE.03' ,
         'ROR1_04_NUMSPECSTOT' =	'MS.MS.count.2R.EE.04' ,
         'ROR2_03_NUMSPECSTOT' =	'MS.MS.count.2R.EE.05' ,
         'ROR2_04_NUMSPECSTOT' =	'MS.MS.count.2R.EE.06' ,
         'PRICKLE1_03_NUMSPECSTOT' =	'MS.MS.count.2R.EE.07', 
         'PRICKLE1_04_NUMSPECSTOT' =	'MS.MS.count.2R.EE.08' ,
         'VANGL2_03_NUMSPECSTOT' = 'MS.MS.count.2R.EE.09' ,
         'VANGL2_04_NUMSPECSTOT' = 'MS.MS.count.2R.EE.10' ,
         'DVL3_03_NUMSPECSTOT' = 'MS.MS.count.2R.EE.11' ,
         'DVL3_04_NUMSPECSTOT' = 'MS.MS.count.2R.EE.12' ,
         'CTRL_05_NUMSPECSTOT' = 'MS.MS.count.3R.01' ,
         'CTRL_06_NUMSPECSTOT' = 'MS.MS.count.3R.02' ,
         'ROR1_05_NUMSPECSTOT' = 'MS.MS.count.3R.03' ,
         'ROR1_06_NUMSPECSTOT' = 'MS.MS.count.3R.04' ,
         'ROR2_05_NUMSPECSTOT' = 'MS.MS.count.3R.05' ,
         'ROR2_06_NUMSPECSTOT' = 'MS.MS.count.3R.06' ,
         'PRICKLE1_05_NUMSPECSTOT' =	'MS.MS.count.3R.07', 
         'PRICKLE1_06_NUMSPECSTOT' =	'MS.MS.count.3R.08' ,
         'VANGL2_05_NUMSPECSTOT' = 'MS.MS.count.3R.09' ,
         'VANGL2_06_NUMSPECSTOT' = 'MS.MS.count.3R.10' ,
         'DVL3_05_NUMSPECSTOT' = 'MS.MS.count.3R.11' ,
         'DVL3_06_NUMSPECSTOT' ='MS.MS.count.3R.12' ) 
```

# Prepare REPRINT output

```{r prepare-reprint, message=FALSE, warning=FALSE}
# Add additional row based on REPRINT requirements
vec <- colnames(data)
vec[1:3] <- "--"
vec <- gsub("(\\_\\d+\\_[A-Z]+)", "", vec)
vec <- gsub("CTRL", "C", vec)

# join the row with dataframe
joined <- rbind (vec, data)

# write txt file
write.table(joined, here('outputs', '03_REPRINT-analysis', '03_REPRINT-data-input.txt'), sep = '\t', row.names = FALSE, quote = FALSE)

```

# REPRINT settings description

Run on 2023-01-02

**Upload data**

* Organism: H.Sapiens
* Experiment type: Proximity Dependent Biotinylation
* Quantitation type: SPC
* File to upload: 05_REPRINT-data-input.txt
* File type: matrix
* File format: tab-separated

**Select controls**

None

**Score interactions**

Empirical Fold Change Score (--default settings)

* Primary Score: user controls ; combining replicates - average ; nr of virtual controls - 10,
* Secondary Score: all controls ; combining replicates - geometric ; nr of virtual controls - 3,

Probabilistic SAINT Score

* SaintExpress
* user controls ; combining replicates - average; nr of virtual controls - 10,
* SAINT options: n-burn 2000, n-iter 4000, LowMode 0, MinFold 1, Normalize 1

Interaction Specificity Score and Enrichment cut offs were not used.

Job name: Wnt/PCP BioID

**Results**

Saved to `analysis/outputs/03_REPRINT-analysis`

* Raw SAINT results: 16544.output
* ProHits input: 16544_prohitsViz_input -> 16544_results.txt
* Cytoscape input: 16544.gml
* Matrix format: 16544_matrix.tab -> saved as .txt