---
title: "04: Dotplots"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

# Libraries

```{r libraries, message=FALSE, warning=FALSE}
library(dplyr)
library(here)
library(DEP)
library(ggplot2)
library(tidyverse)
library(ggdendro)
library(cowplot)
library(ggtree)
```

# Data input

```{r data-input, message=FALSE, warning=FALSE}
data <- read.delim(here('data', '16544_matrix.txt'))
```

Prepare a folder for the outputs
```{r outputs-folder, message=FALSE, warning=FALSE}
#dir.create(here('outputs', '03_REPRINT-analysis'))
if (!dir.exists(here('outputs', '04_Dotplots'))){
  dir.create(here('outputs', '04_Dotplots'))
}
```

Same filters as were used previously for dotplots applied:
- protein is retained if SAINT probability (SP) was in at least one bait >=0.8 (primary score)
- secondary score is used only for visualization

```{r data-filtering, message=FALSE, warning=FALSE}
data.filtered <- data %>%
  filter(if_any(ends_with("SP"), ~ . >= 0.8))
```

# Transform data

Create the same metrics as used in the Prohits visualization tool.

```{r data-transformation, message=FALSE, warning=FALSE}
# unique preys
markers <- data.filtered$GENE %>% unique()

data.filtered <- data.filtered %>%
    select(GENE, contains("SP"), contains("_AVG_"))
  
# compute relative abundance
data.filtered <- data.filtered %>%
  rowwise() %>%
  mutate(DVL3_rel = DVL3_AVG_SPC/  sum(across(DVL3_AVG_SPC:VANGL2_AVG_SPC)) * 100) %>%
  mutate(ROR1_rel = ROR1_AVG_SPC/  sum(across(DVL3_AVG_SPC:VANGL2_AVG_SPC)) * 100) %>%
  mutate(ROR2_rel = ROR2_AVG_SPC/  sum(across(DVL3_AVG_SPC:VANGL2_AVG_SPC)) * 100) %>%
  mutate(PRICKLE1_rel = PRICKLE1_AVG_SPC/  sum(across(DVL3_AVG_SPC:VANGL2_AVG_SPC)) * 100) %>%
  mutate(VANGL2_rel = VANGL2_AVG_SPC/  sum(across(DVL3_AVG_SPC:VANGL2_AVG_SPC)) * 100) 

# create long table
data.long <- pivot_longer(data.filtered, -GENE, names_to = c("protein", ".value"), names_sep = "_")
```

# Dotplots

Comment: in the Rmd file they look weird, it is preferred to plot them in R script and save manually from the Viewer

```{r dotplots, message=FALSE, warning=FALSE}

# dendrogram
mat <- data.filtered %>%
  select(contains("_AVG_"))

row.names(mat) <- data.filtered$GENE
clust <- hclust(dist(mat %>% as.matrix(), method = 'canberra'), method = 'ward.D') # hclust with distance matrix
# for clustering used default option for ProHITS Dotplot

# ddgram <- as.dendrogram(clust) # create dendrogram
# ggtree_plot <- ggtree::ggtree(ddgram)

# plot the dotplot
data.long %>%
  mutate(PreyF = factor(GENE, levels = clust$labels[clust$order])) %>% 
  filter(rel > 0, AVG > 0) %>% 
  mutate(SaintProbability = case_when(
    SP >= 0.9 ~ "SP >= 0.9",
    SP >= 0.7 ~ "SP >= 0.7",
    TRUE ~ "ns"
  )) %>%
  ggplot(aes(protein, PreyF))+
  geom_point(aes(fill = AVG, color = SaintProbability, size = rel, stroke = SP) , shape = 21 ) +  
 # scale_fill_continuous(trans = 'reverse') +
  scale_fill_gradientn(colours = rev(viridis::mako(20)), limits = c(0,50), oob = scales::squish, name = 'Avg abundance') +
  scale_size_continuous(name = 'Relative abundance in %') +
  scale_color_manual(values = c( '#bbdffb', "#56B1F7", "#132B43"))+
  scale_x_discrete(name="Baits") +
  scale_y_discrete(name="Preys") +
  coord_fixed(ratio = 0.8) +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# plot_grid(ggtree_plot, NULL,p, nrow = 1, rel_widths = c(0.5,-0.05, 1), align = 'h')

```

# Cluster-wise dotplots

```{r cluster-heatmap, message=FALSE, warning=FALSE}

# Read the cluster data
load(here('outputs',  '01_DEP-processing', '01_DEP-heatmap-clusters.RData'))

# Filter only relevant columns (protein + cluster assignment)
data.heatmap.filt <- data.heatmap %>% select(protein, k)
```

For each cluster, grep the proteins from the heatmap and order them alphabetically
```{r cluster-dotplots, message=FALSE, warning=FALSE}
# map clusters onto data.long
data.long <- left_join(data.long, data.heatmap.filt, by = c("GENE" = "protein"))
data.long$k[is.na(data.long$k)] <- "not.significant"

# for (i in unique(data.long$k)) {
# p <- data.long %>%
#   filter(k == 1) %>%
#   mutate(PreyF = factor(sort(GENE))) %>% 
#   filter(rel > 0, AVG > 0) %>% 
#   mutate(SaintProbability = case_when(
#     SP >= 0.9 ~ "SP >= 0.9",
#     SP >= 0.7 ~ "SP >= 0.7",
#     TRUE ~ "ns"
#   )) %>%
#   ggplot(aes(protein, PreyF))+
#   geom_point(aes(fill = AVG, color = SaintProbability, size = rel, stroke = SP) , shape = 21 ) +  
#  # scale_fill_continuous(trans = 'reverse') +
#   scale_fill_gradientn(colours = rev(viridis::mako(20)), limits = c(0,50), oob = scales::squish, name = 'Avg abundance') +
#   scale_size_continuous(name = 'Relative abundance in %') +
#   scale_color_manual(values = c( '#bbdffb', "#56B1F7", "#132B43"))+
#   scale_x_discrete(name="Baits") +
#   scale_y_discrete(name="Preys") +
#   coord_fixed(ratio = 0.8) +
#   theme_classic()+
#   theme(axis.text.x = element_text(angle = 45, hjust = 1))+
#   labs(title = 1)
# 
# svg(paste("dotplot_cluster", i, '.svg'))
# print(p)
# dev.off()
# }

# filter out the non-significant ones and change the order based on the clusters
# saved manually with height 2600, same width

data.long %>%
#  filter(k != "not.significant") %>%
  mutate(k = factor(k)) %>%
  arrange(k) %>%
  group_by(k) %>%
  mutate(PreyF = factor(sort(GENE))) %>% 
  filter(rel > 0, AVG > 0) %>%
  mutate(SaintProbability = case_when(
    SP >= 0.9 ~ "SP >= 0.9",
    SP >= 0.7 ~ "SP >= 0.7",
    TRUE ~ "ns"
  )) %>%
  ggplot(aes(protein, PreyF))+
  geom_point(aes(fill = AVG, color = SaintProbability, size = rel, stroke = SP) , shape = 21 ) +
  # scale_fill_continuous(trans = 'reverse') +
  scale_fill_gradientn(colours = rev(viridis::mako(20)), limits = c(0,50), oob = scales::squish, name = 'Avg abundance') +
  scale_size_continuous(name = 'Relative abundance in %') +
  scale_color_manual(values = c( '#bbdffb', "#56B1F7", "#132B43"))+
  scale_x_discrete(name="Baits") +
  scale_y_discrete(name="Preys") +
  coord_fixed(ratio = 0.8) +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
```

Create input file for the metascape based on the clusters

```{r cluster-metascape, message=FALSE, warning=FALSE}
# metascape.input <- data.frame("#Cluster" = c("1", "2", "3", "4", "5"),
#                               "Genes" = c(paste(as.character(data.heatmap$protein[data.heatmap$k == 1]),collapse=", ",sep=""),
#                                           paste(as.character(data.heatmap$protein[data.heatmap$k == 2]),collapse=", ",sep=""),
#                                           paste(as.character(data.heatmap$protein[data.heatmap$k == 3]),collapse=", ",sep=""),
#                                           paste(as.character(data.heatmap$protein[data.heatmap$k == 4]),collapse=", ",sep=""),
#                                           paste(as.character(data.heatmap$protein[data.heatmap$k == 5]),collapse=", ",sep="")))
# 
# write.csv(metascape.input, here('outputs', '05_REPRINT-analysis', '05_metascape-input.csv'), row.names = FALSE)
```

Metascape run on 2023-10-03

Settings: multiple genes analysis, manually reformatted to input (columns = clusters + genes)
express analysis run
