---
title: "04: Dotplots"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

# Setup 

Remove all objects from the workspace
```{r clear-workspace, message=FALSE, warning=FALSE}
rm(list = ls())
```

Prepare a folder for the outputs
```{r outputs-folder, message=FALSE, warning=FALSE}
if (!dir.exists(here('outputs', '04_Dotplots'))){
  dir.create(here('outputs', '04_Dotplots'))
}
```

Libraries required
```{r libraries, message=FALSE, warning=FALSE}
library(dplyr)
library(here)
library(DEP)
library(ggplot2)
library(tidyverse)
library(ggdendro)
library(cowplot)
library(ggtree)
```

# Data input

```{r data-input, message=FALSE, warning=FALSE}
data <- read.delim(here('data', '16544_matrix.txt'))
```

Same filters as were used previously for dotplots applied:
- protein is retained if SAINT probability (SP) was in at least one bait >=0.8 (primary score)
- secondary score is used only for visualization

```{r data-filtering, message=FALSE, warning=FALSE}
data.filtered <- data %>%
  filter(if_any(ends_with("SP"), ~ . >= 0.8))
```

# Transform data

Create the same metrics as used in the Prohits visualization tool.

```{r data-transformation, message=FALSE, warning=FALSE}
# unique preys
markers <- data.filtered$GENE %>% unique()

data.filtered <- data.filtered %>%
    select(GENE, contains("SP"), contains("_AVG_"))
  
# compute relative abundance
data.filtered <- data.filtered %>%
  rowwise() %>%
  mutate(DVL3_rel = DVL3_AVG_SPC/  sum(across(DVL3_AVG_SPC:VANGL2_AVG_SPC)) * 100) %>%
  mutate(ROR1_rel = ROR1_AVG_SPC/  sum(across(DVL3_AVG_SPC:VANGL2_AVG_SPC)) * 100) %>%
  mutate(ROR2_rel = ROR2_AVG_SPC/  sum(across(DVL3_AVG_SPC:VANGL2_AVG_SPC)) * 100) %>%
  mutate(PRICKLE1_rel = PRICKLE1_AVG_SPC/  sum(across(DVL3_AVG_SPC:VANGL2_AVG_SPC)) * 100) %>%
  mutate(VANGL2_rel = VANGL2_AVG_SPC/  sum(across(DVL3_AVG_SPC:VANGL2_AVG_SPC)) * 100) 

# create long table
data.long <- pivot_longer(data.filtered, -GENE, names_to = c("protein", ".value"), names_sep = "_")
```

# Dotplots

*Comment: in the Rmd file they look weird, it is preferred to plot them separately in R script and save manually from the Viewer with adjusted size settings*

Dotplots visualizing the results from REPRINT analysis of BioID data.

The same settings for visualization are used as when done by the ProHits software.

Primary threshold is SP >= 0.9, secondary threshold is SP >= 0.7.

```{r dotplots, message=FALSE, warning=FALSE}
# dendrogram
mat <- data.filtered %>%
  select(contains("_AVG_"))

row.names(mat) <- data.filtered$GENE
clust <- hclust(dist(mat %>% as.matrix(), method = 'canberra'), method = 'ward.D') # hclust with distance matrix
# for clustering used default option for ProHITS Dotplot

# ddgram <- as.dendrogram(clust) # create dendrogram
# ggtree_plot <- ggtree::ggtree(ddgram)

# plot the dotplot

#svg(here('outputs', '04_Dotplots', '04_dotplot_BioID.svg'))
data.long %>%
  mutate(PreyF = factor(GENE, levels = clust$labels[clust$order])) %>% 
  filter(rel > 0, AVG > 0) %>% 
  mutate(SaintProbability = case_when(
    SP >= 0.9 ~ "SP >= 0.9",
    SP >= 0.7 ~ "SP >= 0.7",
    TRUE ~ "ns"
  )) %>%
  ggplot(aes(protein, PreyF))+
  geom_point(aes(fill = AVG, color = SaintProbability, size = rel, stroke = SP) , shape = 21 ) +  
 # scale_fill_continuous(trans = 'reverse') +
  scale_fill_gradientn(colours = rev(viridis::mako(20)), limits = c(0,50), oob = scales::squish, name = 'Avg abundance') +
  scale_size_continuous(name = 'Relative abundance in %') +
  scale_color_manual(values = c( '#bbdffb', "#56B1F7", "#132B43"))+
  scale_x_discrete(name="Baits") +
  scale_y_discrete(name="Preys") +
  coord_fixed(ratio = 0.8) +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#dev.off()

# plot_grid(ggtree_plot, NULL,p, nrow = 1, rel_widths = c(0.5,-0.05, 1), align = 'h')

```

Dotplots for the **non-significant proteins**

```{r cluster-dotplots-nonsignificant, message=FALSE, warning=FALSE}
#svg(here('outputs', '04_Dotplots', '04_dotplot_not-sign.svg'))
data.long %>%
  filter(k == "not.significant") %>%
  mutate(k = factor(k)) %>%
  arrange(k) %>%
  group_by(k) %>%
  mutate(PreyF = factor(sort(GENE))) %>% 
  filter(rel > 0, AVG > 0) %>%
  mutate(SaintProbability = case_when(
    SP >= 0.9 ~ "SP >= 0.9",
    SP >= 0.7 ~ "SP >= 0.7",
    TRUE ~ "ns"
  )) %>%
  ggplot(aes(protein, PreyF))+
  geom_point(aes(fill = AVG, color = SaintProbability, size = rel, stroke = SP) , shape = 21 ) +
  # scale_fill_continuous(trans = 'reverse') +
  scale_fill_gradientn(colours = rev(viridis::mako(20)), limits = c(0,50), oob = scales::squish, name = 'Avg abundance') +
  scale_size_continuous(name = 'Relative abundance in %') +
  scale_color_manual(values = c( '#bbdffb', "#56B1F7", "#132B43"))+
  scale_x_discrete(name="Baits") +
  scale_y_discrete(name="Preys") +
  coord_fixed(ratio = 0.8) +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

dev.off()
```

# Dotplots for heatmap clusters

Dotplots for individual **clusters in the heatmap (#C1-#C5)**.

```{r cluster-heatmap, message=FALSE, warning=FALSE}

# Read the cluster data
load(here('outputs',  '01_DEP-processing', '01_DEP-heatmap-clusters.RData'))

# Filter only relevant columns (protein + cluster assignment)
data.heatmap.filt <- data.heatmap %>% select(protein, k)
```

For each cluster, grep the proteins from the heatmap and order them alphabetically
```{r cluster-dotplots, message=FALSE, warning=FALSE}
# map clusters onto data.long
data.long <- left_join(data.long, data.heatmap.filt, by = c("GENE" = "protein"))
data.long$k[is.na(data.long$k)] <- "not.significant"

#svg(here('outputs', '04_Dotplots', '04_dotplot_clusters.svg'))
data.long %>%
#  filter(k != "not.significant") %>%
  mutate(k = factor(k)) %>%
  arrange(k) %>%
  group_by(k) %>%
  mutate(PreyF = factor(sort(GENE))) %>% 
  filter(rel > 0, AVG > 0) %>%
  mutate(SaintProbability = case_when(
    SP >= 0.9 ~ "SP >= 0.9",
    SP >= 0.7 ~ "SP >= 0.7",
    TRUE ~ "ns"
  )) %>%
  ggplot(aes(protein, PreyF))+
  geom_point(aes(fill = AVG, color = SaintProbability, size = rel, stroke = SP) , shape = 21 ) +
  # scale_fill_continuous(trans = 'reverse') +
  scale_fill_gradientn(colours = rev(viridis::mako(20)), limits = c(0,50), oob = scales::squish, name = 'Avg abundance') +
  scale_size_continuous(name = 'Relative abundance in %') +
  scale_color_manual(values = c( '#bbdffb', "#56B1F7", "#132B43"))+
  scale_x_discrete(name="Baits") +
  scale_y_discrete(name="Preys") +
  coord_fixed(ratio = 0.8) +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#dev.off()
```


